---
title: "Final Project"
author: "Jacob Bowling"
format:
  pdf:
    fontsize: 11pt
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r}
# Load core tidyverse tools for data cleaning, summarizing, and plotting.
library(tidyverse)

# Load lubridate so game dates can be parsed from character strings.
library(lubridate)

# Load GGally for the scatterplot matrix used in the multivariate normality check.
library(GGally)

# Load factoextra for possible PCA/clustering support and visualization utilities.
library(factoextra)

# Load knitr and kableExtra to create formatted tables in the PDF.
library(knitr)
library(kableExtra)

# Load randomForest for the bagging and random forest prediction models.
library(randomForest)

# Load patchwork to combine multiple plots into one figure.
library(patchwork)

# Read in the pitcher-start dataset from the local CSV file.
mlb <- read_csv("C:/Users/jacob/OneDrive/Documents/mlb_2025_starts_with_streamer_category.csv", show_col_types = FALSE)

# Clean and prepare the main dataset for analysis.
mlb_clean <- mlb %>%
  mutate(
    # Convert the game date into an actual Date object.
    game_date = mdy(game_date),

    # Treat pitcher, opponent, and ballpark as categorical variables.
    pitcher_name = factor(pitcher_name),
    opponent = factor(opponent),
    ballpark = factor(ballpark),

    # Convert streamer category to an ordered factor based on start confidence.
    streamer_category = factor(
      streamer_category,
      levels = c(
        "Do Not Start",
        "Questionable Start",
        "Probably Start",
        "Auto Start"
      )
    )
  ) %>%
  # Sort each pitcher's starts chronologically before calculating rest/start order.
  arrange(pitcher_name, game_date) %>%
  group_by(pitcher_name) %>%
  mutate(
    # Calculate days since the pitcher's previous start.
    days_rest = as.numeric(game_date - lag(game_date)),

    # Create a within-pitcher start counter.
    start_number = row_number() + 1
  ) %>%
  ungroup()

# Keep only rows with observed fantasy points for EDA.
eda_df <- mlb_clean %>%
  filter(!is.na(fantasy_points))

# Create the numeric dataset used for PCA.
pca_df <- mlb_clean %>%
  select(
    last_start_fp,
    season_avg_fp_prior,
    days_rest,
    start_number
  ) %>%
  drop_na()

# at one point this was different from pca_df and it was easier to keep it like this than to rewrite everything
# Create the numeric dataset used for clustering.
cluster_df <- mlb_clean %>% 
  select(
    last_start_fp,
    season_avg_fp_prior,
    days_rest,
    start_number
  ) %>%
  drop_na()

# Create the modeling dataset for random forest prediction.
rf_df <- mlb_clean %>%
  select(
    fantasy_points,
    last_start_fp,
    season_avg_fp_prior,
    days_rest,
    start_number,
    opponent,
    ballpark,
    streamer_category
  ) %>%
  # Remove rows with missing values in the response or predictors.
  drop_na() %>%
  mutate(
    # Make sure categorical predictors are treated as factors by the model.
    across(c(opponent, ballpark, streamer_category), as.factor)
  )
```

# Data Description

This dataset contains MLB pitcher-start data from the 2025 season. Each row represents one pitcher start from the 2025 MLB season. The response variable is fantasy points for that start, while the main predictors include PitcherList streamer category, prior fantasy performance, opponent, ballpark, days of rest, and start number. Since the analysis focuses on predicting single-start fantasy outcomes using information available before the start, each pitcher’s first start of the season was excluded because prior-start information was not yet available.

# EDA

## Basic Data Preview

```{r}
# Summarize the main fantasy-point outcome and key numeric predictors.
eda_summary <- eda_df %>%
  summarise(
    # Number of pitcher starts included in the EDA dataset.
    n = n(),

    # Basic distribution of fantasy-point outcomes.
    mean_fp = mean(fantasy_points),
    sd_fp = sd(fantasy_points),
    min_fp = min(fantasy_points),
    max_fp = max(fantasy_points),

    # Average values for the main pre-start numeric predictors.
    mean_last_start = mean(last_start_fp, na.rm = TRUE),
    mean_prior_avg = mean(season_avg_fp_prior, na.rm = TRUE),
    mean_days_rest = mean(days_rest, na.rm = TRUE)
  ) %>%
  # Round numeric values so the table is easier to read.
  mutate(across(where(is.numeric), ~round(., 3)))

# Print the summary statistics table.
kable(
  eda_summary,
  booktabs = TRUE,
  align = rep("c", ncol(eda_summary))
) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Summary Statistics" = ncol(eda_summary)))
```

Fantasy point outcomes vary widely, with a large range and high std relative to the mean.

## Correlation

```{r}
# Compute the correlation matrix for the numeric predictors used in PCA.
corr_table <- pca_df %>%
  cor(use = "complete.obs") %>%
  round(3)

# Print the correlation matrix as a formatted table.
kable(
  corr_table,
  booktabs = TRUE,
  align = rep("c", ncol(corr_table))
) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Correlation Matrix for Numeric Predictors" = 5))
```

The strongest relationship is between last-start fantasy points and prior season average fantasy points, indicating a moderate association. All other correlations are weak, suggesting that rest and timing variables capture different aspects of pitcher usage. Overall, the low correlations indicate limited multicollinearity and support using these predictors in later analyses.

## PitcherList Streamer Category

```{r}
# Summarize fantasy-point outcomes within each PitcherList streamer category.
streamer_summary <- eda_df %>%
  group_by(streamer_category) %>%
  summarise(
    # Count the number of starts in each category.
    n = n(),

    # Summarize the distribution of fantasy points by category.
    mean_fp = mean(fantasy_points),
    sd_fp = sd(fantasy_points),
    min_fp = min(fantasy_points),
    max_fp = max(fantasy_points),
    median_fp = median(fantasy_points),

    # Include prior performance to compare category strength with past production.
    mean_prior_avg = mean(season_avg_fp_prior, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Round numeric values for table readability.
  mutate(across(where(is.numeric), ~round(., 3)))

# Print the category summary table.
kable(
  streamer_summary,
  booktabs = TRUE,
  align = rep("c", ncol(streamer_summary))
) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Fantasy Points by Streamer Category" = ncol(streamer_summary)))
```

```{r}
#| fig-width: 6
#| fig-height: 2.5
#| fig-align: center

# Create a boxplot comparing fantasy points across streamer categories.
ggplot(eda_df, aes(x = streamer_category, y = fantasy_points)) +
  # Show the spread and median fantasy points within each category.
  geom_boxplot() +
  labs(
    x = "Streamer Category",
    y = "Fantasy Points"
  ) +
  theme_classic()
```

The streamer categories generally move in the expected direction, with Auto Start pitchers having the highest median fantasy points and Do Not Start pitchers having the lowest. (The NA category is likely due to their start being postponed). The large overlap between categories shows that expert classifications provide useful signal but do not fully separate good and bad outcomes.

## Prior Performance

```{r}
#| fig-width: 5
#| fig-height: 2.25
#| fig-align: center

# Plot prior average fantasy points against actual fantasy points for the start.
ggplot(eda_df, aes(x = season_avg_fp_prior, y = fantasy_points)) +
  # Add individual pitcher-start observations.
  geom_point(alpha = 0.4) +
  # Add a linear trend line to show the overall relationship.
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Prior Season Average Fantasy Points",
    y = "Fantasy Points"
  ) +
  theme_classic()
```

Prior season average fantasy points has a positive relationship with actual fantasy points, but the points remain widely scattered. This suggests that prior performance matters, while single-start outcomes still contain substantial randomness.

## Opponent & Ballpark

```{r}
# Manually enter final team records so opponent quality can be shown alongside fantasy outcomes.
team_records <- tribble(
  ~opponent, ~wins, ~losses,
  "MIL", 97, 65,
  "PHI", 96, 66,
  "NYY", 94, 68,
  "TOR", 94, 68,
  "LAD", 93, 69,
  "CHC", 92, 70,
  "SDP", 90, 72,
  "BOS", 89, 73,
  "HOU", 88, 74,
  "SEA", 87, 75,
  "CLE", 86, 76,
  "SFG", 81, 81,
  "ARI", 80, 82,
  "CIN", 83, 79,
  "NYM", 83, 79,
  "MIN", 82, 80,
  "STL", 78, 84,
  "TEX", 82, 80,
  "DET", 79, 83,
  "MIA", 79, 83,
  "ATL", 76, 86,
  "BAL", 75, 87,
  "TBR", 77, 85,
  "KCR", 76, 86,
  "LAA", 68, 94,
  "WSN", 66, 96,
  "OAK", 65, 97,
  "CHW", 60, 102,
  "PIT", 71, 91,
  "COL", 43, 119
) %>%
# Convert wins and losses into a formatted winning percentage.
  mutate(win_pct = scales::percent(wins / (wins + losses),
      accuracy = 0.1))
```

```{r}
#| results: asis

# Summarize fantasy outcomes by opponent.
opponent_summary <- eda_df %>%
  group_by(opponent) %>%
  summarise(
    mean_fp = round(mean(fantasy_points), 2),
    median_fp = round(median(fantasy_points), 0),
    n = n(),
    .groups = "drop"
  ) %>%
  left_join(team_records, by = "opponent") %>%
  select(opponent, win_pct, mean_fp, median_fp, n) %>%
  rename(team = opponent) %>%
  arrange(desc(mean_fp))

# Identify opponents associated with the highest average pitcher fantasy points.
top5_opp <- opponent_summary %>% slice_head(n = 5)
# Identify opponents associated with the lowest average pitcher fantasy points.
bottom5_opp <- opponent_summary %>% slice_tail(n = 5)  %>% arrange(mean_fp)

cat("\\begin{minipage}{0.48\\linewidth}
\\centering
\\textbf{Worst Opponents}\\\\[4pt]
{\\small
")

print(kable(top5_opp, format = "latex", booktabs = TRUE))

cat("}
\\end{minipage}
\\hspace{0.03\\linewidth}
\\begin{minipage}{0.48\\linewidth}
\\centering
\\textbf{Best Opponents}\\\\[4pt]
{\\small
")

print(kable(bottom5_opp, format = "latex", booktabs = TRUE))

cat("}
\\end{minipage}")
```

```{r}
#| results: asis

# Summarize fantasy outcomes by ballpark.
ballpark_summary <- eda_df %>%
  group_by(ballpark) %>%
  summarise(
    mean_fp = round(mean(fantasy_points), 2),
    median_fp = round(median(fantasy_points), 0),
    n = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_fp))

# Identify ballparks associated with the highest average pitcher fantasy points.
top5_park <- ballpark_summary %>% slice_head(n = 5)
# Identify ballparks associated with the lowest average pitcher fantasy points.
bottom5_park <- ballpark_summary %>% slice_tail(n = 5) %>% arrange(mean_fp)

cat("\\begin{minipage}{0.48\\textwidth}
\\centering
\\textbf{Most Pitcher-Friendly Ballparks}\\\\[4pt]
")

print(
  kable(top5_park, format = "latex", booktabs = TRUE)
)

cat("\\end{minipage}
\\hfill
\\begin{minipage}{0.48\\textwidth}
\\centering
\\textbf{Least Pitcher-Friendly Ballparks}\\\\[4pt]
")

print(
  kable(bottom5_park, format = "latex", booktabs = TRUE)
)

cat("\\end{minipage}")
```

Certain opponents/ballparks are clearly better/worse. Interestingly not all of the "worst opponents" were "bad teams" by record.

# PCA

## Assessment of Multivariate Normality

```{r}
#| fig-width: 10
#| fig-height: 4.5

# Create a scatterplot matrix to inspect pairwise relationships and marginal distributions.
p_pairs <- ggpairs(
  pca_df,
  progress = FALSE,
  lower = list(continuous = wrap("points", alpha = 0.35, size = 0.5)),
  upper = list(continuous = wrap("cor", size = 3)),
  diag  = list(continuous = wrap("densityDiag", alpha = 0.4))
)

# Convert the GGally object into a grob so it can be combined with patchwork.
pairs_grob <- GGally::ggmatrix_gtable(p_pairs)
p_pairs_wrapped <- patchwork::wrap_elements(full = pairs_grob)

# Reshape the PCA variables to build separate univariate Q-Q plots.
pca_long <- pca_df %>%
  pivot_longer(everything(), names_to = "var", values_to = "value")

# Create univariate Q-Q plots for each numeric variable.
p_qq_uni <- ggplot(pca_long, aes(sample = value)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~var, scales = "free") +
  labs(title = "Univariate Q-Q Plots")

# Compute the covariance matrix and mean vector for Mahalanobis distances.
S <- cov(pca_df)
xbar <- colMeans(pca_df)
# Calculate squared Mahalanobis distances for the multivariate normality check.
d2 <- mahalanobis(pca_df, center = xbar, cov = S)

# Create theoretical chi-square quantiles and ordered observed distances.
theo <- qchisq(ppoints(length(d2)), df = ncol(pca_df))
obs  <- sort(d2)

# Plot observed Mahalanobis distances against chi-square quantiles.
p_qq_chi <- ggplot(tibble(theo = theo, obs = obs), aes(theo, obs)) +
  geom_point(alpha = 0.7, size = 1) +
  geom_abline(intercept = 0, slope = 1) +
  labs(
    title = expression(chi^2 ~ "Q-Q"),
    x = expression(chi^2 ~ Quantiles),
    y = "Ordered Distances"
  )

# Combine the scatterplot matrix and Q-Q plots into one diagnostic figure.
p_pairs_wrapped / (p_qq_uni | p_qq_chi) +
  plot_layout(heights = c(2, 1))
```

Multivariate normality was assessed using pairwise scatterplots, univariate Q–Q plots, and a $\chi^2$ Q–Q plot. The **scatterplot matrix** shows a moderate positive relationship between last start fantasy points and prior season average fantasy points, while other relationships appear weak. The **univariate Q–Q plots** indicate departures from normality, particularly for days of rest, which is strongly right-skewed, and start number, which also has deviation. The $\chi^2$ **Q–Q plot** exhibits substantial upward deviation in the upper tail, indicating heavier-than-normal tails and potential outliers. The assumption of multivariate normality is not satisfied.

## Principal Component Analysis

```{r}
# Fit PCA using centered and scaled numeric predictors.
pca_fit <- prcomp(pca_df, center = TRUE, scale. = TRUE)
```

Because the variables are measured in different units, they were standardized prior to analysis, as recommended when scale differences may otherwise dominate the first principal component.

## Eigenvalue Table

```{r}
# Extract eigenvalues from the PCA object.
eig_vals <- pca_fit$sdev^2
# Calculate the proportion of variance explained by each component.
prop_var <- eig_vals / sum(eig_vals)
# Calculate cumulative variance explained.
cum_var <- cumsum(prop_var)

# Build a table summarizing eigenvalues and explained variance.
var_table <- tibble(
  PC = paste0("PC", 1:length(eig_vals)),
  Eigenvalue = eig_vals,
  Proportion = prop_var,
  Cumulative = cum_var
)

# Print the eigenvalue/explained-variance table.
kable(var_table, digits = 3)
```

The first three principal components explain 86.5% of the total variation in the four numeric variables. PC1 explains 37.6%, PC2 explains 26.1%, and PC3 explains 22.8%. Since these three components preserve most of the original information, the first three components are retained for interpretation.

## Loadings

```{r}
# Extract the PCA loading matrix.
loadings <- pca_fit$rotation
# Print the loadings for the first three principal components.
kable(loadings[, 1:3], digits = 3)
```

**PC1** has large negative loadings for last_start_fp and season_avg_fp_prior, so it mainly represents prior pitcher performance. More negative PC1 values correspond to stronger recent and season-level performance. **PC2** is characterized by a strong positive loading for start_number and a negative loading for days_rest. This component primarily contrasts later-season starts with more typical rest patterns against earlier or irregularly spaced starts with longer rest intervals. **PC3** has large positive loadings for both days_rest and start_number, indicating that it captures variation related to the timing and spacing of starts. In particular, higher PC3 values correspond to starts that occur later in the season and involve longer rest periods, suggesting less typical scheduling patterns (or return from injury).

## Score Plot

```{r}
#| fig-width: 4
#| fig-height: 2
#| fig-align: center

# Convert PCA scores to a data frame for plotting.
scores <- as.data.frame(pca_fit$x)

# Plot PC1 vs. PC2 and color points by PC3.
ggplot(scores, aes(x = PC1, y = PC2, color = PC3)) +
  # Add one point for each pitcher start.
  geom_point(alpha = 0.7) +
  # Use a continuous color scale for PC3 values.
  scale_color_viridis_c() +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "PC3"
  ) +
  theme_minimal()
```

The score plot displays the first two principal components, with color representing values of the third principal component. Most observations are concentrated in the upper-left region of the plot, indicating that the majority of pitcher starts have relatively similar combinations of prior-performance and timing characteristics, corresponding to lower PC1 values and slightly higher PC2 values. A smaller set of observations extends away from this cluster, particularly toward higher PC1 and lower PC2 values, with lighter colors indicating higher PC3 values. These points represent less typical starts, especially those involving longer rest periods and later-season appearances (probably due to injuries). Overall, the plot suggests continuous variation rather than distinct clusters.

# K-Means Clustering

## Elbow Plot

```{r}
#| fig-width: 8
#| fig-height: 4

# Standardize clustering variables so all variables contribute on the same scale.
cluster_scaled <- scale(cluster_df)

# Set seed so k-means results are reproducible.
set.seed(0827)

# Create a list to store k-means fits for k = 1 through 8.
k_list <- vector("list", 8)

# Fit k-means models across candidate numbers of clusters.
for(k in 1:8){
  k_list[[k]] <- kmeans(cluster_scaled, centers = k, nstart = 50)
}

# Extract total within-cluster sum of squares for the elbow plot.
twcss <- sapply(k_list, function(x) x$tot.withinss)

# Plot total within-cluster SS against the number of clusters.
plot(1:8, twcss,
     type = "b",
     xlab = "Number of Clusters",
     ylab = "Total Within-Cluster SS")
```

The elbow plot shows substantial reductions in within-cluster variation when increasing the number of clusters from one to two and from two to three. A further decrease is observed when moving to four clusters, after which the rate of improvement diminishes. Therefore, a four-cluster solution was selected to capture additional structure in the data while maintaining a relatively simple clustering model.

## Cluster Means

```{r}
# Set seed so the final k-means solution is reproducible.
set.seed(0827)

# Fit the selected four-cluster k-means model.
mlb_kmeans <- kmeans(
  cluster_scaled,
  centers = 4,
  nstart = 50,
  iter.max = 200
)
```

```{r}
# Attach k-means cluster assignments to the original-scale variables and summarize each cluster.
mlb_cluster_means <-
  mlb_clean %>%
  # Select variables used to interpret cluster profiles.
  select(
    fantasy_points,
    last_start_fp,
    season_avg_fp_prior,
    days_rest,
    start_number
  ) %>%
  drop_na() %>%
  # Add the k-means cluster label for each row.
  mutate(cluster = mlb_kmeans$cluster) %>%
  # Average each variable within each cluster.
  group_by(cluster) %>%
  summarise(across(everything(), mean)) %>%
  rename(mean_fp = fantasy_points) %>%
  mutate(across(-cluster, ~round(., 3)))

# Print the k-means cluster means table.
kable(mlb_cluster_means, booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("K-Means Cluster Means" = 6))
```

This identifies distinct groups of pitcher starts based on prior performance and timing characteristics. **Cluster 3** has the highest average fantasy points and is characterized by strong recent performance and high prior season averages, indicating consistently high-performing pitchers. **Cluster 1** represents more typical starts, with values near the overall mean across all variables. **Clusters 2 and 4** are associated with lower average fantasy point outcomes. **Cluster 2** consists of starts with weaker recent and prior performance but otherwise typical rest patterns. **Cluster 4** is distinguished by extremely high days of rest, suggesting irregular scheduling (potentially due to returning from injury), and also exhibits lower fantasy point production. Clusters were created from pre-start characteristics and then evaluated using actual fantasy outcomes. While the clusters capture differences in prior performance and scheduling patterns, they interestingly do not show as strong of a separation in terms of fantasy point outcomes as one would expect.

# Hierarchical Clustering

## Linkage Choice

```{r}
#| fig-width: 6
#| fig-height: 2.5

# Set seed so the sampled subset is reproducible.
set.seed(0827)

# Sample 500 observations to make the hierarchical clustering dendrogram readable.
sample_idx <- sample(nrow(cluster_scaled), 500)
# Subset the scaled clustering data to the sampled rows.
cluster_sample <- cluster_scaled[sample_idx, ]

# Compute pairwise distances among sampled observations.
dist_cluster <- dist(cluster_sample)

# Fit hierarchical clustering using three linkage methods for comparison.
hc_single   <- hclust(dist_cluster, method = "single")
hc_complete <- hclust(dist_cluster, method = "complete")
hc_average  <- hclust(dist_cluster, method = "average")

# Display the three dendrograms side by side.
par(mfrow = c(1,3))

plot(hc_single,   main = "Single Linkage",   xlab = "", sub = "")
plot(hc_complete, main = "Complete Linkage", xlab = "", sub = "")
plot(hc_average,  main = "Average Linkage",  xlab = "", sub = "")
```

Hierarchical clustering was performed using single, complete, and average linkage methods. The single linkage dendrogram exhibits strong chaining behavior, with observations merging gradually and little clear separation between groups. Average linkage provides some improvement, but clusters remain relatively compressed with limited vertical separation. Complete linkage produces more distinct cluster structure, with larger vertical gaps between merges and clearer separation among groups. Therefore, complete linkage was selected for further analysis.

## Cluster Means

```{r}
# Cut the complete-linkage dendrogram into four clusters.
hc_clusters <- cutree(hc_complete, 4)

# Attach hierarchical cluster assignments and summarize original-scale means.
mlb_hc_means <-
  mlb_clean %>%
  select(
    fantasy_points,
    last_start_fp,
    season_avg_fp_prior,
    days_rest,
    start_number
  ) %>%
  drop_na() %>%
  # Keep the same sampled rows used to fit the dendrogram.
  slice(sample_idx) %>%
  # Add the hierarchical cluster label.
  mutate(cluster = hc_clusters) %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean)) %>%
  rename(mean_fp = fantasy_points) %>%
  mutate(across(-cluster, ~round(., 3)))
```

```{r}
# Print the hierarchical cluster means table.
kable(mlb_hc_means,
      booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Hierarchical Cluster Means" = 6))
```

This identifies distinct groups of pitcher starts based on prior performance and timing characteristics. **Cluster 2** stands out with the highest average fantasy point outcomes and reflects consistently strong recent and overall performance. **Cluster 1** represents more typical starts, with values closer to the overall average. **Clusters 3 and 4** are distinguished by extremely high rest periods, indicating irregular scheduling (potentially due to returning from injury). Despite these differences, both clusters do not show substantially different fantasy point outcomes than the others. Clusters were created from pre-start characteristics and then evaluated using actual fantasy outcomes. Again, the clusters capture differences in performance and timing patterns, but they do not show strong separation in terms of fantasy point outcomes.

## Comparison

The k-means and hierarchical clustering solutions both identify groups of pitcher starts based on prior performance and timing patterns. However, the k-means solution provides a more interpretable separation between groups. In particular, it distinguishes starts with strong prior performance, weaker prior performance, typical starts, and irregular long-rest starts. Because of this k-means is preferred for describing pitcher-start profiles.

# PCA/Clustering Visual

```{r}
#| fig-width: 5
#| fig-height: 2.25
#| fig-align: center

# Add k-means cluster labels to the PCA score data.
cluster_scores <- as.data.frame(pca_fit$x) %>%
  mutate(cluster = factor(mlb_kmeans$cluster))

# Plot the k-means clusters in the PC1-PC2 space.
ggplot(cluster_scores, aes(x = PC1, y = PC2, color = cluster)) +
  # Add one point for each pitcher start.
  geom_point(alpha = 0.5, size = 2.2) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Cluster"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )
```

The k-means clusters are visualized using the first two principal components here. The clusters generally align with the structure identified in the PCA, as observations that are close together in the principal component space tend to belong to the same cluster. One cluster is clearly separated from the others, while the remaining clusters are more tightly grouped and show some overlap. The visualization reinforces that both PCA and clustering are capturing similar underlying patterns in the data (even though the separation between groups is not especially strong).

# Random Forest

```{r}
# Set seed so the train/test split is reproducible.
set.seed(0827)

# Randomly select 70% of rows for the training set.
train <- sample(1:nrow(rf_df), floor(0.7 * nrow(rf_df)))

# Create training data from the sampled row indices.
train_data <- rf_df[train, ]
# Create testing data from the remaining rows.
test_data  <- rf_df[-train, ]
# Store the observed test-set fantasy points for error calculations.
test_y <- test_data$fantasy_points

# Count the number of predictors for use in bagging/random forest mtry settings.
p <- ncol(train_data) - 1
```

```{r}
# Set seed so the random forest results are reproducible.
set.seed(0827)

# Fit a bagging model by using all predictors at each split.
bag_fit <- randomForest(
  fantasy_points ~ .,
  data = train_data,
  mtry = p,
  ntree = 500,
  importance = TRUE
)

# Predict fantasy points on the test set using the bagging model.
bag_pred <- predict(bag_fit, newdata = test_data)
# Compute the test-set MSE for the bagging model.
bag_mse <- mean((bag_pred - test_y)^2)

# Define candidate mtry values for random forest tuning.
mtry_vals <- 1:(p - 1)
# Create storage for test-set MSE values.
rf_mse_vals <- numeric(length(mtry_vals))
# Create storage for the fitted random forest models.
rf_fits <- vector("list", length(mtry_vals))

# Fit and evaluate a random forest for each candidate mtry value.
for (i in seq_along(mtry_vals)) {
  rf_fits[[i]] <- randomForest(
    fantasy_points ~ .,
    data = train_data,
    mtry = mtry_vals[i],
    ntree = 500,
    importance = TRUE
  )

  # Predict on the test set for the current model.
  rf_pred_i <- predict(rf_fits[[i]], newdata = test_data)
  # Store the test-set MSE for the current mtry value.
  rf_mse_vals[i] <- mean((rf_pred_i - test_y)^2)
}
```

## Results

```{r}
# Combine bagging and random forest test errors into one comparison table.
results_tbl <- tibble(
  Model = c("Bagging", rep("Random Forest", length(mtry_vals))),
  mtry = c(p, mtry_vals),
  ntree = 500,
  Test_MSE = c(bag_mse, rf_mse_vals)
) %>%
  # Sort so the best-performing model appears first.
  arrange(Test_MSE) %>%
  mutate(Test_MSE = round(Test_MSE, 3))

# Print the model comparison table.
kable(results_tbl, booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Model Comparison" = 4))
```

The random forest with mtry = 1 produced the lowest test-set mean squared error, making it the best-performing model among those considered. As mtry increases, the test-set MSE increases slightly, indicating that using fewer predictors at each split leads to marginally better performance. Bagging, which uses all predictors, resulted in the highest test-set MSE. However, the differences in test error across models are relatively small, suggesting that while limiting predictors improves performance, the overall predictive gains are modest.

## Variable Importance

```{r}
# Extract the best random forest mtry value from the model comparison table.
best_mtry <- results_tbl %>%
  filter(Model == "Random Forest") %>%
  slice(1) %>%
  pull(mtry)

# Pull the fitted random forest model corresponding to the best mtry value.
rf_final <- rf_fits[[which(mtry_vals == best_mtry)]]

# Convert the random forest importance output into a data frame.
var_imp_tbl <- as.data.frame(importance(rf_final))
# Preserve predictor names as their own column.
var_imp_tbl$Predictor <- rownames(var_imp_tbl)

# Format and sort the variable importance table.
var_imp_tbl <- var_imp_tbl %>%
  as_tibble() %>%
  select(Predictor, `%IncMSE`, IncNodePurity) %>%
  arrange(desc(`%IncMSE`)) %>%
  mutate(
    `%IncMSE` = round(`%IncMSE`, 3),
    IncNodePurity = round(IncNodePurity, 3)
  )

# Print the variable importance table.
kable(var_imp_tbl, booktabs = TRUE) %>%
  kable_styling(latex_options = "hold_position", font_size = 11) %>%
  add_header_above(c("Variable Importance" = 3))
```

Variable importance was assessed using percent increase in mean squared error. The most important predictor was streamer category. Prior season average fantasy points was the second most important predictor, suggesting that overall previous performance was more useful than performance in only the most recent start. Opponent and ballpark also contributed to predictive accuracy, indicating that matchup and environment matter, but their effects were smaller than streamer category and prior performance. Days of rest and start number were the least important predictors, suggesting that scheduling variables added less predictive value in this model.

## Predicted vs. Observed Plot

```{r}
#| fig-width: 6
#| fig-height: 4

# Generate test-set predictions from the final selected random forest model.
final_pred <- predict(rf_final, newdata = test_data)

# Plot predicted fantasy points against observed fantasy points.
plot(
  final_pred, test_y,
  xlab = "Predicted Fantasy Points",
  ylab = "Observed Fantasy Points"
)
# Add a 45-degree reference line showing perfect prediction.
abline(0, 1)
```

The predicted vs. observed plot shows a positive relationship, but predictions are compressed into a narrow range while actual values vary widely. This indicates regression to the mean, with high outcomes underpredicted and low outcomes overpredicted. The model captures some signal but leaves substantial unexplained variation.

# Conclusion

This analysis shows how pre-start characteristics relate to fantasy point outcomes using PCA, clustering, and random forests. The exploratory analysis showed moderate relationships between prior performance measures and fantasy outcomes, along with substantial variability across observations. PCA showed that most variation in the data could be explained by prior performance and timing patterns, while clustering methods identified groups of starts with similar profiles. However, both k-means and hierarchical clustering showed only limited separation in actual fantasy outcomes. The random forest model provided some predictive improvement and identified streamer category and prior performance as the most important predictors. Despite this, the predicted versus observed results showed substantial unexplained variation, indicating that even with multiple variables and nonlinear modeling, performance remains difficult to predict. These results suggest that while pre-start information and expert classifications provide useful signal, they have limited ability to fully explain or predict fantasy performance, highlighting the inherent variability in pitcher outcomes.

# References

Pendles, J. (2026). *2025 MLB Streaming Statistics*. Kaggle.\
     https://www.kaggle.com/datasets/jabrwocky/2025-mlb-streaming-statistics
